Temporal ordering and registration of images in studies of developmental dynamics

### 2014-06-06 image_analysis_paper/si_information.tex
Added everything before working on submission

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets (and half of the median of the pairwise distances between data points for the third data set).

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$,
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
%
As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate.
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%%
%%YGK just as a favor to me, tell me how consistent the three values are -- 
%%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$

@@

To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
this portion of the sphere is small enough so that curvature effects are not prominent in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

@@

%
\begin{figure*}[t]
{\bf Item C in the figure below is our main observation.}

\raisebox{7.6cm}{\sffamily{A}}
\includegraphics[width=8cm]{raw_data1}
\hfill
\raisebox{7.6cm}{\sffamily{B}}
\includegraphics[width=8cm]{FBsPCA_evals}

\raisebox{5.7cm}{\sffamily{C}}
\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
%
I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
%
The results are shown above.
%
However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
%
(A) Images used for FBsPCA. (B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients. There is no visible spectral gap after the first component, as would be expected for this particular data set. (C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness. Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form. }
\end{figure*}

>>>>

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances between data points for the first two data sets (and half of the median of the squared pairwise distances between data points for the third data set).

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$,
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
%
As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate.
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%%
%%YGK just as a favor to me, tell me how consistent the three values are -- 
%%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
\begin{equation} \label{eq:first_embed_vdm}
\max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. 
\end{equation}

\begin{table}[t]
%\caption{Outline of algorithm used to temporally order images}
\setlength{\fboxrule}{2pt}
\framebox[\columnwidth]{
\begin{minipage}{0.9\columnwidth}
{\bf Outline of algorithm used to register and order images}
\begin{enumerate}
\item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
Retain both the optimal pairwise alignments (translations and rotations) as well as the minimum distances obtained when pairs are optimally aligned.
%
\item Compute the corresponding pairwise three-dimensional rotations from the translations and rotations, using \eqref{eq:angle_relations}, \eqref{eq:three_rot}, and \eqref{eq:total_rot}.
%
\item Select $\epsilon$ for use in the diffusion maps kernel. 
We often begin with using the median of the pairwise distances as $\epsilon$, and then adjust as needed.
\item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$, which are computed in step 2.
%
\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
%
\item Estimate the optimal three-dimensional rotation for each image: stack the first three eigenvectors $\phi_1, \phi_2, \phi_3$ into a $3m \times 3$ matrix, and then divide this matrix into $m$ $3 \times 3$ blocks, $\hat{R}_1, \dots, \hat{R}_m$. Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}. If most of the rotations are improper ($det(R) = -1$), flip the sign of one of the eigenvectors.
\item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
\item For each image $i$, compute the optimal translation and rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding translation and rotation, described in \eqref{eq:angle_invert}. 
\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $4 \le k \le 6$.
\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
\end{enumerate} 
\end{minipage}
}
\end{table}

@@

To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
this portion of the sphere is small enough so that curvature effects are not prominent in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation} \label{eq:three_rot}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

@@

%
%\begin{figure*}[t]
%{\bf Item C in the figure below is our main observation.}
%
%\raisebox{7.6cm}{\sffamily{A}}
%\includegraphics[width=8cm]{raw_data1}
%\hfill
%\raisebox{7.6cm}{\sffamily{B}}
%\includegraphics[width=8cm]{FBsPCA_evals}
%
%\raisebox{5.7cm}{\sffamily{C}}
%\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
%\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
%%
%I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
%%
%The results are shown above.
%%
%However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
%%
%(A) Images used for FBsPCA. (B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients. There is no visible spectral gap after the first component, as would be expected for this particular data set. (C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness. Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form. }
%\end{figure*}
